Bayesian Sparsification of $\cplx$-valued networks

### 2019-11-07 paper/paper.tex
cplx-gaussian comment edits
moved notation from local-rep-trick, simplified approximation descriptionn

%notation
We first clarify some notation: $M^{\top}$ is the matrix transpose, $\conj{M}$ is elementwise
complex conjugation, and $M^{\hop} = (\conj{M})^\top$ denotes Hermitian conjugate.

@@

the characteristic function is
$$
z \sim \mathcal{N}^{\cplx}_m(z \vert \mu, \Gamma, C)
  \Leftrightarrow
  \mathbb{E}_{z} e^{j \Re{(t^{\hop} z)}}
  = \exp{\{
    j \Re{(t^{\hop} \mu)}
    - \tfrac14 \bigl(
      t^{\hop} \Gamma t + \Re{(t^{\hop} C \conj{t})}
    \bigr)
  \}}
  \,, $$

Not every $\real^{2m}$ Gaussian corresponds to a $\cplx^m$ Gaussian vecor,
but when it does the covariance and relation can be recovered through
$$
\Sigma
  = \begin{pmatrix}
    \Sigma_{rr} & \Sigma_{ri} \\ 
    \Sigma_{ir} & \Sigma_{ii} 
  \end{pmatrix}
  \, \Leftrightarrow
  \,
  \begin{cases}
    \Gamma
      = \Sigma_{rr} + \Sigma_{ii} + j (\Sigma_{ir} - \Sigma_{ri})
    \,, \\
    C = \Sigma_{rr} - \Sigma_{ii} + j (\Sigma_{ir} + \Sigma_{ri}) 
  \end{cases}
  $$

This implies that $
  \Sigma_{ir}
  % = -\Sigma_{ri}
  = -\Sigma_{ir}^\top
$ and $\Sigma_{rr} = \Sigma_{ii}$, i.e. real and imaginary component of the same random element
are uncorrelated, since skew-symmetric matrices necessarily have zero-diagonal.

If $M \in \real^{m\times m}$, $M \succeq 0$ and $M^\top = M$, then
$$
  (u - j v)^\top M (u + j v)
    = u^\top M u + v^\top M v + j (u^\top M v - v^\top M u)
    % = u^\top M u + v^\top M v + j (u^\top M v - u^\top M^\top v)
    = u^\top M u + v^\top M v \geq 0
\,. $$
if $M^\top = - M$, then
$$
  u^\top M v - v^\top M u - j (u^\top M u + v^\top M v)
  % = u^\top M v - u^\top M^\top v - j (u^\top M u + v^\top M^\top v)
  = u^\top M v + u^\top M v - j (u^\top M u - v^\top M v)
  \,. $$

%
% The intuition behind them is more apparent in the complex univariate case ($m=1$):
% here $\Gamma = \sigma^2 \geq 0$, and $C$ can be represented as $\sigma^2 \rho \in \cplx$ with $
%   \lvert \rho \rvert \leq 1
% $. Therefore $\sigma^2$ affects the overall ``concentration'' of the complex random variable,
% whereas $\rho$ is responsible for the strength and ``direction'' of the linear relation between
% the real and imaginary parts, \cite{lapidoth_capacity_2003}.
% $$
% \frac{\sigma^2}2
%   \begin{pmatrix}
%     1 + \Re{\rho} & \Im{\rho} \\
%     \Im{\rho} & 1 - \Re{\rho}
%   \end{pmatrix}
%   = \frac{\sigma^2}2 I
%   + \frac{\sigma^2 \lvert \rho \rvert}2 
%   \begin{pmatrix}
%     \cos\theta & \sin\theta \\
%     \sin\theta & - \cos\theta
%   \end{pmatrix}
% $$

To develop $\cplx$-variational dropout we assume a fully factorized posterior approximation
distribution on the layers' parameters: $\omega = (W, b)$, where a $\cplx$-valued $W$ collects
all weight matrices of linear layers and kernels of convolutional layers, and $b$ represents
all bias term combined. We use a fully factorized Dirac family on the biases, and mean-field
proper $\cplx$-Gaussian variational approximation for the weights: $
  q(W) = \prod_{lij} q(W^{l}_{ij})
$ with each parameter being an independent univariate proper complex Gaussian
% (we omit the subscript $m=1$ for brevity)
\begin{equation}  \label{eq:c-gauss-vi}
  W_{ij}
    \sim \mathcal{N}^{\cplx}(\mu_{ij}, \sigma^2_{ij}, 0)
  \,.
\end{equation}
Thus we would like to sparsify the weights in the network and, since the bias terms constitute
a meagre share of network's parameters and are rarely dropped out or pruned, we use Bayesan
MAP estimates for them.

@@

Consider a $\cplx$-linear layer: $
  W \in \cplx^{n\times m}
$ is a random $\cplx$-Gaussian matrix with independent elements $
  w_{ij} \sim \mathcal{N}^{\cplx}\bigl(
    w_{ij}
    \big \vert
    \mu_{ij}, \sigma^2_{ij}, C_{ij}
  \bigr)
$ with $\mu_{ij}\in cplx$, $\sigma^2_{ij} \geq 0$, $C_{ij}\in \cplx$ and $
  \lvert C_{ij} \rvert^2 \leq \sigma^2_{ij}
$. We begin by observing that the variational approximation of $q(W)$ is a multivariate
$\cplx$-Gaussian: $
  \vec{W}
    \sim \mathcal{N}^{\cplx}_{[n\times m]} \bigl(
      \vec{\mu}, \diag{\vec{\Sigma}}, \diag{\vec{C}}
    \bigr)
$, where $[n\times m]$ denotes lexicographically flattened dimension and $\diag{v}$ is a
diagonal square matrix with $v_i$ on the main diagonal. Here we collected variances and means
into matrices $\Sigma$ and $\mu$, respectively.
%
Now, for any $x \in \cplx^m$ and $b\in \cplx^n$ we have
$$
W x + b
  % = I_n W x + b
  = (I_n \otimes x^\top) \vec{W} + b
  \,, $$
where $\vec{\cdot}$ denotes flattening in lexicographic order of indices (row-major), and
$\otimes$ denotes the Kronecker product, for which we have the identity $
  \vec{P Q R} = (P \otimes R^\top) \vec{Q}
$, \cite{petersen_matrix_2012}.
% It follows from the definition of the Kronecker product
% of $P$ and $R^\top$ and the row-major order vectorization.
%
Structure of the matrices, involved in this expression, and the identity $
  (P \otimes R) (C \otimes S) = P Q \otimes R S
$, \cite{petersen_matrix_2012}, imply
\begin{align*}
  (I \otimes x^\top) \diag{\vec{\Sigma}} (I \otimes x^\top)^\hop
    % &= \sum_{ij} (I \otimes x^\top) e_i \otimes e_j
    %   \sigma^2_{ij} (e_i \otimes e_j)^\top (I \otimes x^\top)^\hop
    % \\
    &
    = \sum_{ij} (e_i \otimes x^\top e_j)
      \sigma^2_{ij} (e_i \otimes x^\top e_j)^\hop
    \notag \\
    % &= \sum_{ij} (e_i \otimes x_j) \sigma^2_{ij} (e_i^\top \otimes \conj{x}_j)
    % \\
    &
    = \sum_i (e_i e_i^\top) \sum_j \sigma^2_{ij} x_j \conj{x}_j
    \,,
\end{align*}
where $e_i$ is the $i$-th real unit vector of dimensionality conforming to the expression it
is involved with. In a similar fashion we obtain the expression for the relation matrix of
the transformation:
\begin{equation*}
  (I \otimes x^\top) \diag{\vec{C}} (I \otimes x^\top)^\top
    % = \sum_{ij} (e_i \otimes x_j) C_{ij} (e_i \otimes x_j)^\top
    = \sum_i e_i e_i^\top \sum_j C_{ij} x_j^2
    \,.
\end{equation*}
The term $
  \sum_i e_i e_i^\top (\ldots)
$ means that the covariance and relation matrices of $y = Wx + b$ are diagonal.

>>>>

%notation
We first clarify some notation: $M^{\top}$ is the matrix transpose, $\conj{M}$ is elementwise
complex conjugation, and $M^{\hop} = (\conj{M})^\top$ denotes Hermitian conjugate. By $e_i$
we denote the $i$-th real unit vector of dimensionality \textbf{conforming} to the matrix-vector
expression it is used in. $\vec{M}$ denotes row-major flattening of $M$ into a vector, i.e.
in lexicographic order of its indices. $\otimes$ denotes the Kronecker product, for which we
note the following identities $
  \vec{P Q R} = (P \otimes R^\top) \vec{Q}
$, and $
  (P \otimes R) (C \otimes S) = P Q \otimes R S
$, \cite{petersen_matrix_2012}.
% It follows from the definition of the Kronecker product
% of $P$ and $R^\top$ and the row-major order vectorization.

@@

%
% The roles of $\Gamma$ and $C$ is more evident in univariate case: here $
%   \Gamma = \sigma^2 \geq 0
% $, and $C = \sigma^2 \rho \in \cplx$ with $
%   \lvert \rho \rvert \leq 1
% $. $\sigma^2$ determines the ``dispersion'', whereas $\rho$ is determines the
% ``strength'' and ``direction'' of the linear relation between real and imaginary
% parts, \cite{lapidoth_capacity_2003}.
% $$
% \frac{\sigma^2}2
%     1 + \Re{\rho} & \Im{\rho} \\
%     \Im{\rho} & 1 - \Re{\rho}
%   = \frac{\sigma^2}2 
%         1 + \lvert \rho \rvert \cos\theta
%           & \lvert \rho \rvert \sin\theta \\
%         \lvert \rho \rvert \sin\theta
%           & 1 - \lvert \rho \rvert \cos\theta
%   \,. $$

In $\cplx$-variational dropout we use mean-field proper $\cplx$-Gaussian variational
approximation. Specifically, parameters $W \in \cplx^{n\times m}$ of a dense layer follow
\begin{equation}  \label{eq:c-gauss-vi}
  % (we omit the subscript $m=1$ for brevity)
  W \sim q(W) = \prod_{ij}
    \mathcal{N}^{\cplx}(W_{ij} \mid \mu_{ij}, \sigma^2_{ij}, 0)
  \,,
\end{equation}
and we use analogous independent proper $\cplx$-Gaussian approximation for convolutional
kernels. As for bias terms, we propose to use their MAP estimates, i.e. assume an independent
fully factorized Dirac approximation.

@@

Consider a $\cplx$-linear layer: $
  W \in \cplx^{n\times m}
$ is a random $\cplx$-Gaussian matrix with independent elements $
  w_{ij} \sim \mathcal{N}^{\cplx}\bigl(
    w_{ij}
    \big \vert
    \mu_{ij}, \sigma^2_{ij}, C_{ij}
  \bigr)
$ with $\mu_{ij}\in cplx$, $\sigma^2_{ij} \geq 0$, $C_{ij}\in \cplx$ and $
  \lvert C_{ij} \rvert^2 \leq \sigma^2_{ij}
$. We begin by observing that the variational approximation of $q(W)$ is a multivariate
$\cplx$-Gaussian: $
  \vec{W}
    \sim \mathcal{N}^{\cplx}_{[n\times m]} \bigl(
      \vec{\mu}, \diag{\vec{\Sigma}}, \diag{\vec{C}}
    \bigr)
$, where $[n\times m]$ denotes lexicographically flattened dimension and $\diag{v}$ is a
diagonal square matrix with $v_i$ on the main diagonal. Here we collected variances and means
into matrices $\Sigma$ and $\mu$, respectively.
%
Now, for any $x \in \cplx^m$ and $b\in \cplx^n$ we have
$$
W x + b
  % = I_n W x + b
  = (I_n \otimes x^\top) \vec{W} + b
  \,. $$
Structure of the matrices, involved in this expression, imply
\begin{align*}
  (I \otimes x^\top) \diag{\vec{\Sigma}} (I \otimes x^\top)^\hop
    % &= \sum_{ij} (I \otimes x^\top) e_i \otimes e_j
    %   \sigma^2_{ij} (e_i \otimes e_j)^\top (I \otimes x^\top)^\hop
    % \\
    &
    = \sum_{ij} (e_i \otimes x^\top e_j)
      \sigma^2_{ij} (e_i \otimes x^\top e_j)^\hop
    \notag \\
    % &= \sum_{ij} (e_i \otimes x_j) \sigma^2_{ij} (e_i^\top \otimes \conj{x}_j)
    % \\
    &
    = \sum_i (e_i e_i^\top) \sum_j \sigma^2_{ij} x_j \conj{x}_j
    \,.
\end{align*}
In a similar fashion we obtain the expression for the relation matrix of
the transformation:
\begin{equation*}
  (I \otimes x^\top) \diag{\vec{C}} (I \otimes x^\top)^\top
    % = \sum_{ij} (e_i \otimes x_j) C_{ij} (e_i \otimes x_j)^\top
    = \sum_i e_i e_i^\top \sum_j C_{ij} x_j^2
    \,.
\end{equation*}
The term $
  \sum_i e_i e_i^\top (\ldots)
$ means that the covariance and relation matrices of $y = Wx + b$ are diagonal.
